Transient Corotating Clumps Around Adolescent Low-Mass Stars From Four Years of TESS

### 2023-09-06 paper/ms.tex
Rewrote abstract
Co-authored-by: Lynne Hillenbrand <lah@astro.caltech.edu>

% 250 word max!
\begin{abstract}
  We explore a set of pre-main-sequence stars
  with sub-solar masses, rotation periods of 0.2--2 days, optical variability amplitudes
  of a few percent, and sharp light curve features that repeat for
  hundreds of rotation cycles.  The sharp features are probably caused by
  eclipsing clumps of dust or gas orbiting at the corotation radius.
  We refer to this class of star as ``complex periodic variables'' (CPVs).
  To improve our understanding of these
  enigmatic objects, we conducted a survey for CPVs using TESS
  short-cadence data collected between 2018~July and 2022~September.
  Our search of \nstarssearched\ K and M dwarfs with $T$$<$16 and
  $d$$<$150\,pc yielded \ngoods\ high-quality CPVs, most of which are
  new discoveries.  Our sample includes the brightest
  ($T$$\approx$9.5), closest ($d$$\approx$20\,pc), and oldest
  ($\approx$200\,Myr) known CPVs.  One of the new discoveries, LP
  12-502, stands out for the complexity of its light curve --- with as
  many as eight flux dips per cycle --- and for the detail with which 
  each dip's evolving amplitude and phase can be tracked over
  1{,}500 cycles.  We argue that corotating gas or dust remains the
  most viable explanation for CPVs. The gas or dust is probably
  entrained by the star's magnetic field, and the sharp features could
  result from a multipolar field topology, a hypothesis supported by
  correspondences between the light curves of CPVs and of B stars
  known to have multipolar magnetic fields.
\end{abstract}

@@

While theoretical heritage for understanding magnetic
rotators exists, CPVs have remained mysterious because they have been both
hard to discover and hard to characterize.   They have been hard to discover
because they are rare: CPVs comprise $\approx$1\% of the youngest
$\approx$1\% of M dwarfs \citep{2018AJ....155..196R}.  Out of the
millions of stars monitored by K2 and TESS, about 50 CPVs have been
reported to date
\citep{2016AJ....152..114R,2017AJ....153..152S,2018AJ....155...63S,2019ApJ...876..127Z,2020AJ....160...86B,2021AJ....161...60S,2022AJ....163..144G,2023ApJ...945..114P}.
They have been hard to characterize because many of the known CPVs are
faint; the initial K2 discoveries
\citep{2016AJ....152..114R,2017AJ....153..152S} were M2-M6 dwarfs at
distances $\gtrsim$100\,pc, with optical brightnesses of
$V$$\approx$15.5 to $V$$>$20.  At such magnitudes, high-resolution
time-series spectroscopy is out of reach with current facilities,
despite the potential utility of such observations.

@@

{\sc State-changes reveal dip independence}---The state-changes seen
in cycles 1233-1247 and 299-315 confirm that dips can disappear in
less than one cycle. While such behavior was also noted by
\cite{2017AJ....153..152S}, the data presented here show further that
the dips can be {\it independent} and {\it additive}.  For example,
throughout cycles 1233-1264, there are three sharp dips between phases
of 0 and 0.3 with different amplitudes but similar slopes.  During the
transition, the leading dip nearly disappeared while the other two
dips hardly changed; compare the centermost two panels of
Figure~\ref{fig:lp}.  Evidently, the material or process responsible
for one dip can vary independently of the materials or processes
responsible for other dips.  The state changes during cycles 248-264
and 299-315 support the same conclusion, while also hinting that the
{\it leading} dip of a complex is most prone to disappearing, leaving
the trailing dips unchanged in its wake.

@@

%{\bf AUTHOR CONTRIBUTION STATEMENTS: PLEASE REVISE IF APPROPRIATE)}
LGB conceived the project, performed the dip-counting search, light curve classification, cluster membership, SED, variability, and secondary-period analyses, and wrote the manuscript.
RJ and SR performed the Fourier-based analysis and contributed to light curve classification.
LR cross-examined the light curve classification, and contributed an independent SED analysis.
AD identified the magnetic B star connection.
LAH contributed to project design and to the interpretation of the light curves.
JNW and SR significantly improved the clarity of the manuscript.
G\'AB acquired and maintained the servers used to run the dip-finding pipeline.
%-----
%{\bf JNW, MNG, GRR to confirm coauthorship \&/or to help out on interpretation!}
% Pending!
%HI and AH designed and maintain the architecture used to acquire and process the HIRES spectra.
All authors assisted in manuscript revision.

@@

{\it How many of our CPVs are new?}---Considering only our ``good''
CPVs, 35 of these 50 sources have not yet been reported in the
literature.  Considering only the 13 ``ambiguous'' CPVs, 11 are new to
the literature.  Two sources, TIC~67897871 (EPIC~202724025; RIK-90;
\citealt{2017AJ....153..152S}) and TIC~118769116 (EPIC~247343526;
\citealt{2017AJ....153..152S}), were previously only known to be CPVs
due to K2 observations.  Both periods appear to have remained
identical, within measurement precision.  The shapes have of course changed.
These two objects, combined with PTFO 8-8695, provide evidence
that CPV variability persists for at least decade timescales.

>>>>

% 250 word max!
\begin{abstract}
  %
  Complex periodic variables (CPVs) are stars that exhibit sharp and
  periodic optical light curves.  Previous work has shown that such
  stars tend to be disk-free pre-main-sequence M dwarfs with rotation
  periods of 0.2--2 days.  To advance understanding of these enigmatic
  objects, we performed a blind CPV search using TESS 2-minute data of
  \nstarssearched\ K and M dwarfs with $T$$<$16 and $d$$<$150\,pc.  We
  found \ngoods\ high-quality CPVs.  The new objects include the
  brightest ($T$$\approx$9.5), closest ($d$$\approx$20\,pc), and
  oldest ($\approx$200\,Myr) CPVs known.  All objects but one turned
  out to be likely members of stellar associations, which enabled
  age-dating.  One exceptional new object, LP 12-502, showed up to
  eight dips per cycle; the dips at times coexisted with slightly
  different periods, and the variability of the dips suggests that
  they are additive and independent.  Broadly, our search confirms
  that CPVs are mostly young ($\lesssim$150\,Myr) and low-mass
  ($\lesssim$0.4\,$M_\odot$).  The flux dips that embody the class
  have lifetimes of $\approx$100 cycles, and the dips collapse in
  response to stellar flares once every few months.  Corotating gas or
  dust clumps remain the only viable explanation.  The gas or dust is
  probably entrained by the star's magnetic field, and the sharp
  features could result from a multipolar field topology, a hypothesis
  supported by correspondences between the light curves of CPVs and of
  B stars known to have multipolar magnetic fields.
  %
\end{abstract}

@@

While theoretical heritage for understanding magnetic
rotators exists, CPVs have remained mysterious because they have been both
hard to discover and hard to characterize.   They have been hard to discover
because they are rare: CPVs comprise $\approx$1\% of the youngest
$\approx$1\% of M dwarfs \citep{2018AJ....155..196R}.  Out of the
millions of stars monitored by K2 and TESS, about 70 CPVs have been
reported to date
\citep{2016AJ....152..114R,2017AJ....153..152S,2018AJ....155...63S,2019ApJ...876..127Z,2020AJ....160...86B,2021AJ....161...60S,2022AJ....163..144G,2023ApJ...945..114P}.
They have been hard to characterize because many of the known CPVs are
faint; the initial K2 discoveries
\citep{2016AJ....152..114R,2017AJ....153..152S} were M2-M6 dwarfs at
distances $\gtrsim$100\,pc, with optical brightnesses of
$V$$\approx$15.5 to $V$$>$20.  At such magnitudes, high-resolution
time-series spectroscopy is out of reach with current facilities,
despite the potential utility of such observations.

@@

{\sc State-changes reveal dip independence}---The state-changes seen
in cycles 261, 309, and 1241 confirm that dips can disappear in
less than one cycle. While such behavior was also noted by
\cite{2017AJ....153..152S}, the data presented here show further that
the dips can be {\it independent} and {\it additive}.  For example,
throughout cycles 1233-1264, there are three sharp dips between phases
of 0 and 0.3 with different amplitudes but similar slopes.  During the
transition, the leading dip nearly disappeared while the other two
dips hardly changed; compare the centermost two panels of
Figure~\ref{fig:lp}.  Evidently, the material or process responsible
for one dip can vary independently of the materials or processes
responsible for other dips.  The state changes during cycles 261
and 309 support the same conclusion, while also hinting that the
{\it leading} dip of a complex is most prone to disappearing, leaving
the trailing dips unchanged in its wake.

@@

%{\bf AUTHOR CONTRIBUTION STATEMENTS: PLEASE REVISE IF APPROPRIATE)}
LGB conceived the project, performed the dip-counting search, light curve classification, cluster membership, SED, variability, and secondary-period analyses, and wrote the manuscript.
RJ and SR performed the Fourier-based analysis and contributed to light curve classification.
LR cross-examined the light curve classification, and contributed an independent SED analysis.
AD identified the magnetic B star connection.
LAH contributed to project design and to the interpretation of the light curves.
JNW, SR, and LAH significantly improved the clarity of the manuscript.
G\'AB acquired and maintained the servers used to run the dip-finding pipeline.
%-----
%{\bf JNW, MNG, GRR to confirm coauthorship \&/or to help out on interpretation!}
% Pending!
%HI and AH designed and maintain the architecture used to acquire and process the HIRES spectra.
All authors assisted in manuscript revision.

@@

{\it How many of our CPVs are new?}---Considering only our ``good''
CPVs, 35 of these 50 sources have not yet been reported in the
literature.  Considering only the 13 ``ambiguous'' CPVs, 11 are new to
the literature.  Two sources, TIC~67897871 (EPIC~202724025; RIK-90;
\citealt{2017AJ....153..152S}) and TIC~118769116 (EPIC~247343526;
\citealt{2017AJ....153..152S}), were previously only known to be CPVs
due to K2 observations.  Both periods appear to have remained
identical, within measurement precision.  The shapes have of course changed.
These two objects, combined with PTFO 8-8695, provide evidence
that CPV variability persists over timescales of at least decades.
